Reanalysis Step 3: Propensity Score Analysis with Twang

### 2025-08-19 reanalysis-2-multiple_imputation.qmd
refactor to be more quarto friendly

# Print initial imputation model summary
message("Total variables in dataset: ", ncol(df_mice))
message("Variables to be imputed: ", sum(mice_dry$nmis > 0))
message("Total missing values: ", sum(mice_dry$nmis))
message("Missing data by variable:")
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))

print(miss_vars)

method_table <- data.frame(
  Variable = names(mice_dry$method),
  Method = mice_dry$method,
  stringsAsFactors = FALSE
) |>
  filter(Method != "")

message("Default imputation methods:")
print(method_table)

@@

# Print final imputation model configuration
message("FINAL IMPUTATION MODEL")

@@

message("Imputation methods by variable:")
print(final_methods)

# Show which variables will be imputed
vars_to_impute <- names(custom_methods)[custom_methods != ""]
message("Variables that WILL be imputed (", length(vars_to_impute), " total):")
print(vars_to_impute)

# Show predictor matrix summary
message("Predictor matrix summary:")
message("- Each variable uses ", sum(pred_matrix[1,]), " predictors on average")
message("- Variables used as predictors:")
predictors <- names(which(colSums(pred_matrix) > 0))
print(predictors)

# Print model parameters
message("Imputation Parameters:")
message("Mode: ", analysis_mode)
message("Number of imputations (m): ", n_imputations)
message("Max iterations per imputation (maxit): ", n_iterations)
message(paste("Random seed: ", seed))
```

@@

```{r run-imputation}
# Run mice imputation
# Note: Parameters are set based on analysis_mode in setup
message("Starting multiple imputation...")
message("Mode: ", analysis_mode)
message("Imputations: ", n_imputations)
message("Iterations: ", n_iterations)

if (analysis_mode == "test") {
  message("This should take a few minutes...")
} else {
  message("This may take 15-30 minutes or more...")
}

@@

# Apply transformation functions to recreate derived variables
message("Recreating derived variables for each imputed dataset...")

@@

message("Number of incomplete cases in PS model variables per imputed dataset:")
print(unlist(completeness_check))
```

@@

```{r save-imputed}
saveRDS(mice_results, file.path(reanalysis_data_dir, "mice_results.rds"))
saveRDS(imputed_datasets, file.path(reanalysis_data_dir, "imputed_datasets.rds"))
saveRDS(imputed_datasets[[1]], file.path(reanalysis_data_dir, "imputed_dataset_1.rds"))
message("Saved ", n_imputations, " imputed datasets to: ", reanalysis_data_dir)
```

@@

imputed_vars <- names(mice_results$nmis)[mice_results$nmis > 0]
message("Variables that were imputed:")
print(imputed_vars)
```

>>>>

# Imputation model summary
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  filter(N_Missing > 0) |>
  arrange(desc(N_Missing))
```

### Initial Imputation Model Summary

**Total variables in dataset:** `r ncol(df_mice)`  
**Variables to be imputed:** `r sum(mice_dry$nmis > 0)`  
**Total missing values:** `r sum(mice_dry$nmis)`

**Missing data by variable:**
```{r missing-vars-table}
kable(miss_vars)

method_table <- data.frame(
  Variable = names(mice_dry$method),
  Method = mice_dry$method,
  stringsAsFactors = FALSE
) |>
  filter(Method != "")
```

**Default imputation methods:**
```{r method-table}
kable(method_table)

@@

```

### Final Imputation Model Configuration

```{r final-model-config}

@@

vars_to_impute <- names(custom_methods)[custom_methods != ""]
```

**Imputation methods by variable:**
```{r final-methods-table}
kable(final_methods)
```

**Variables that WILL be imputed** (`r length(vars_to_impute)` total):
```{r vars-to-impute}
print(vars_to_impute)

# Show predictor matrix summary
avg_predictors <- sum(pred_matrix[1,])
predictors <- names(which(colSums(pred_matrix) > 0))
```

### Predictor Matrix Summary

Each variable uses an average of **`r avg_predictors`** predictors.

**Variables used as predictors:**
```{r predictor-list}
print(predictors)
```

### Imputation Parameters

```{r imputation-params}
imp_params <- data.frame(
  Parameter = c("Mode", "Number of imputations (m)", "Max iterations (maxit)", "Random seed"),
  Value = c(analysis_mode, n_imputations, n_iterations, seed)
)
kable(imp_params)
```

@@

```{r run-imputation}
# Run mice imputation
# Note: Parameters are set based on analysis_mode in setup
time_estimate <- if(analysis_mode == "test") "a few minutes" else "15-30 minutes or more"
```

**Starting multiple imputation...**

- Mode: `r analysis_mode`
- Imputations: `r n_imputations`
- Iterations: `r n_iterations`
- Estimated time: `r time_estimate`

```{r mice-run}

@@

# Apply transformation functions to recreate derived variables

@@

```

### Completeness Check

**Number of incomplete cases in PS model variables per imputed dataset:**
```{r completeness-results}
print(unlist(completeness_check))
```

@@

```{r save-imputed}
saveRDS(mice_results, file.path(reanalysis_data_dir, "mice_results.rds"))
saveRDS(imputed_datasets, file.path(reanalysis_data_dir, "imputed_datasets.rds"))
saveRDS(imputed_datasets[[1]], file.path(reanalysis_data_dir, "imputed_dataset_1.rds"))
```

**Saved `r n_imputations` imputed datasets to:** `r reanalysis_data_dir`

```{r empty-chunk}
```

@@

imputed_vars <- names(mice_results$nmis)[mice_results$nmis > 0]
```

**Variables that were imputed:**
```{r imputed-vars-list}
print(imputed_vars)
```

### 2025-08-19 reanalysis-3-propensity_scores.qmd
refactor to be more quarto friendly

# Additional packages
library(twang)
library(ggplot2)
library(gridExtra)  # For arranging multiple plots
library(knitr)

@@

message("Analysis mode: ", analysis_mode)
message("Will show diagnostics for up to ", max_imputations_for_diagnostics, " imputations")
message("Using ", n_trees, " trees for GBM")
```

@@

message("Loaded ", n_imputations, " imputed datasets")

@@

message("Propensity score formula:")
print(ps_formula_enhanced)

@@

@@

# Run twang for each imputed dataset
message("Calculating propensity scores using Twang GBM...")
message("Parameters:")
message("  - n.trees: ", n_trees)
message("  - interaction.depth: 3")
message("  - shrinkage: 0.01")
message("  - estimand: ATT")
message("  - stop.method: es.mean (effect size only)")

for (i in 1:n_imputations) {
  message("Processing imputation ", i, " of ", n_imputations, "...")

@@

  message("  - Trees used (es.mean): ", twang_model$desc$es.mean$n.trees)
  message("  - Weight range: [", round(min(weights), 3), ", ", round(max(weights), 3), "]")
}

message("Twang propensity score calculation complete")
```

@@

  if (i <= max_imputations_for_diagnostics) {
    message("Balance summary for imputation ", i, ":")
    print(summary(twang_model))
  }
}

# Combine all summaries
balance_summary_all <- do.call(rbind, balance_summaries)

message("Weight Summary Across All Imputations:")
kable(balance_summary_all, 
      caption = "Weight distribution by imputation",
      digits = 3)

@@

message("Average Weights Across Imputations:")
kable(avg_weights,
      caption = "Average weight statistics across all imputations",
      digits = 3)

if (avg_weights$Max_Weight_Overall > 10) {
  message("Warning: Maximum weight > 10, indicating potential instability")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability")
} else {
  message("Weights appear stable (max < 5)")
}

message("Note: Balance statistics are shown in the summary output above.")
message("The twang package optimizes to minimize effect sizes (es.mean).")
```

@@

message("Generating twang diagnostic plots for ", n_to_show, " imputations...")

# For each selected imputation, show twang-specific diagnostics
for (i in imps_to_show) {
  message("Diagnostics for Imputation ", i, ":")

@@

  tryCatch({
    plot(twang_model)
  }, error = function(e) {
    message("  Standard plots not available, generating custom diagnostics...")

@@

# Plot PS distribution for first few imputations
for (i in imps_to_show) {
  message("Propensity score distributions for imputation ", i)

@@

  # Also show weight summary
  message("  Weight summary:")
  message("    Control: mean = ", round(mean(weights[!treated]), 3), 
          ", max = ", round(max(weights[!treated]), 3))
  message("    Treated: mean = ", round(mean(weights[treated]), 3),
          " (should be 1.0 for ATT)")
}
```

@@

message("Twang propensity score results saved to: ", 
        file.path(reanalysis_data_dir, "ps_results_twang.rds"))
```

@@

```{r summary}
message("Analysis Summary:")
message("Method: Generalized Boosted Models (GBM) via twang")
message("Imputations processed: ", n_imputations)
message("Analysis mode: ", analysis_mode)
message("Trees used: ", n_trees)
message("Estimand: ATT (Average Treatment Effect on the Treated)")
message("Stop method: es.mean (effect size minimization)")

# Enhanced formula summary
message("Enhanced formula included:")
message("  - DR severity (person_dr)")
message("  - Treatment type")
message("  - All standard covariates from original analysis")

# Weight summary  
message("Weight Statistics (averaged across imputations):")
message("  - Mean weight (treated): ", round(avg_weights$Avg_Weight_Treated, 3))
message("  - Mean weight (control): ", round(avg_weights$Avg_Weight_Control, 3))
message("  - Maximum weight: ", round(avg_weights$Max_Weight_Overall, 2))

# Weight stability assessment
if (avg_weights$Max_Weight_Overall > 10) {
  message("  Warning: Maximum weight > 10, indicating potential instability")
  message("  Consider increasing n.trees or adjusting twang parameters")
} else if (avg_weights$Max_Weight_Overall > 5) {
  message("Note: Maximum weight > 5, monitor for stability in outcome analysis")
} else {
  message("Weights appear stable (max < 5)")
}

message("Balance Assessment:")
message("  See summary output above for detailed balance statistics.")
message("  The twang package optimizes es.mean to achieve covariate balance.")
```

>>>>

library(twang)
library(ggplot2)
library(gridExtra)
library(knitr)

@@

```

**Analysis mode:** `r analysis_mode`  
**Will show diagnostics for up to** `r max_imputations_for_diagnostics` **imputations**  
**Using** `r n_trees` **trees for GBM**

@@

```

**Loaded** `r n_imputations` **imputed datasets**

```{r continue-load}

@@

```

**Propensity score formula:**
```{r print-formula}
print(ps_formula_enhanced)

@@

### Model Parameters
- **Method:** Twang GBM
- **n.trees:** Set based on analysis mode (1000 for test, 3000 for final)
- **interaction.depth:** 3
- **shrinkage:** 0.01
- **estimand:** ATT
- **stop.method:** es.mean (effect size only)

@@

for (i in 1:n_imputations) {
  cat("Processing imputation", i, "of", n_imputations, "...\n")

@@

  cat("  - Trees used (es.mean):", twang_model$desc$es.mean$n.trees, "\n")
  cat("  - Weight range: [", round(min(weights), 3), ",", round(max(weights), 3), "]\n")
}

cat("\nTwang propensity score calculation complete\n")
```

@@

  if (i <= max_imputations_for_diagnostics) {
    cat("\nBalance summary for imputation", i, ":\n")
    print(summary(twang_model))
  }
}

balance_summary_all <- do.call(rbind, balance_summaries)

```

### Weight Summary Across All Imputations

```{r weight-summary-table}
kable(balance_summary_all, 
      caption = "Weight distribution by imputation",
      digits = 3)

@@

```

### Average Weights Across Imputations

```{r avg-weights-table}
kable(avg_weights,
      caption = "Average weight statistics across all imputations",
      digits = 3)

```

### Weight Stability Check

```{r weight-stability-check}
if (avg_weights$Max_Weight_Overall > 10) {
  print("Warning: Maximum weight > 10, indicating potential instability")
} else if (avg_weights$Max_Weight_Overall > 5) {
  print("Note: Maximum weight > 5, monitor for stability")
} else {
  print("Weights appear stable (max < 5)")
}
```

@@

cat("\nGenerating twang diagnostic plots for", n_to_show, "imputations...\n")

# For each selected imputation, show twang-specific diagnostics
for (i in imps_to_show) {
  cat("\nDiagnostics for Imputation", i, ":\n")

@@

  tryCatch({
    plot(twang_model)
  }, error = function(e) {
    cat("  Standard plots not available, generating custom diagnostics...\n")

@@

# Plot PS distribution for first few imputations
for (i in imps_to_show) {
  cat("\nPropensity score distributions for imputation", i, "\n")

@@

  cat("  Weight summary:\n")
  cat("    Control: mean =", round(mean(weights[!treated]), 3), 
      ", max =", round(max(weights[!treated]), 3), "\n")
  cat("    Treated: mean =", round(mean(weights[treated]), 3),
      "(should be 1.0 for ATT)\n")
}
```

@@

```

**Twang propensity score results saved to:** `r file.path(reanalysis_data_dir, "ps_results_twang.rds")`

@@

### Analysis Configuration

**Method:** Generalized Boosted Models (GBM) via twang  
**Estimand:** ATT (Average Treatment Effect on the Treated)  
**Stop Method:** es.mean (effect size minimization)

```{r summary-config}
config_summary <- data.frame(
  Parameter = c("Analysis Mode", "Imputations Processed", "Trees per Model"),
  Value = c(analysis_mode, n_imputations, n_trees)
)

kable(config_summary, caption = "Run Configuration")
```

### Formula Components

The enhanced propensity score formula includes:
- DR severity (person_dr)
- Treatment type variables
- All standard covariates from original analysis

```{r summary-formula}
print(ps_formula)
```

### Weight Statistics Summary

```{r summary-weights}
weight_stats <- data.frame(
  Statistic = c("Mean Weight (Treated)", "Mean Weight (Control)", "Maximum Weight"),
  Value = c(round(avg_weights$Avg_Weight_Treated, 3),
           round(avg_weights$Avg_Weight_Control, 3),
           round(avg_weights$Max_Weight_Overall, 2))
)

kable(weight_stats, caption = "Weight Statistics (Averaged Across Imputations)")
```

#### Weight Stability Assessment

```{r weight-stability}
if (avg_weights$Max_Weight_Overall > 10) {
  print("⚠ WARNING: Maximum weight > 10, indicating potential instability")
  print("Consider increasing n.trees or adjusting twang parameters")
} else if (avg_weights$Max_Weight_Overall > 5) {
  print("📊 NOTE: Maximum weight > 5, monitor for stability in outcome analysis")
} else {
  print("✓ GOOD: Weights appear stable (max < 5)")
}
```

### Balance Achievement

The twang package optimizes es.mean to achieve covariate balance. Detailed balance statistics are shown in the diagnostic outputs above. Results have been saved for use in outcome analysis.
